---
title: "Exploratory Data Analysis: Attendance Data"
output: html_notebook
---

## 1. Data Munging
```{r set-up, include=FALSE}
library(tidyverse)
library(silounloadr)
library(janitor)
library(lubridate)
library(naniar)

options(gargle_oauth_email = "mberrien@kippchicago.org")
bigrquery::bq_auth(path = "kipp-chicago-silo-2-aa786970aefd.json")
theme_set(theme_kipp_light())
```


```{r parameters}
SY_1920 <- silounloadr::calc_academic_year(ymd("2019-08-19"), format = "firstyear")

# Output: 2019
CURRENT_FIRST_YEAR <- calc_academic_year(lubridate::today(),
                                         format = "first_year")

# Output: 29
PS_TERMID_1920 <- calc_ps_termid(CURRENT_FIRST_YEAR)

PS_YEARID_1920 <-
  silounloadr::calc_ps_termid(SY_1920) %>%
  str_extract("\\d{2}") %>%
  as.integer()

FIRST_DAY_SCHOOL_1920 <- terms %>%
  filter(id == PS_TERMID_1920) %>% 
  select(firstday)
  
LAST_DAY_SCHOOL_1920 <- terms %>%
  filter(id == PS_TERMID_1920) %>% 
  select(lastday)
```

The "student_daily_attendance" table is produced by joining the `attendance`, `attendance_code` and `students` tables from powerschools. The end result is a table that contains all students who have enrolled in KIPP Chicago and taken classes from 2003 to today. Each row is one day for a specific student. The row notes if the student is enrolled and if they were present, absent or tardy.
```{r}
student_daily_attendance <- 
  
  # this table incldes only days where students where not first marked as present
  attendance %>%
  
  # This table adds the attendance codes (tardy, absent, suspension, etc.)
  left_join(attendance_code,
            by = c("attendance_codeid" = "id")) %>%
  
  # this table ensures that each student has a row for each day they were enrolled in school. 
  # because attendance table only records "special circumstances", membership table is needed
  # to figure out all days students were present. 
  right_join(membership, 
            by = c("studentid" = "studentid", 
                   "att_date" = "date")) %>%
  
  # Identify whether att_code is enrolled, present, absent, or tardy for each student by day
  mutate(enrolled0 = 1,
         enrolled = if_else(att_code == "D" & !is.na(att_code), 0, enrolled0),
         present0 = ifelse(is.na(att_code) | att_code == "", 1, 0),
         present1 = ifelse(att_code %in%  c("A", "S"), 0, present0),
         present2 = ifelse(att_code == "H", 0.5, present1),
         present3 = ifelse(att_code %in% c("T", "E", "L", "I"), 1, present2),
         present = ifelse(is.na(present2), 1, present3),
         absent = (1 - present)*enrolled,
         tardy = ifelse(att_code %in% "T", 1, 0)) %>%
  
  # Add additional student info
  left_join(students,
            by = c("studentid" = "studentid")
            ) %>%
  arrange(yearid, 
          student_number, 
          att_date) %>%
  
  # filter out students who were DNA (students who signed up but never came to school)
  filter(exitcode != "99") %>%
  
  # create human readable school year abbrevations
  mutate(sy = case_when(yearid == "13" ~ "03-04", 
                      yearid == "14" ~ "04-05", 
                      yearid == "15" ~ "06-06", 
                      yearid == "16" ~ "06-07",
                      yearid == "17" ~ "07-08", 
                      yearid == "18" ~ "08-09",
                      yearid == "19" ~ "09-10", 
                      yearid == "20" ~ "10-11",
                      yearid == "21" ~ "11-12", 
                      yearid == "22" ~ "12-13",
                      yearid == "23" ~ "13-14", 
                      yearid == "24" ~ "14-15",
                      yearid == "25" ~ "15-16",
                      yearid == "26" ~ "16-17", 
                      yearid == "27" ~ "17-18",
                      yearid == "28" ~ "18-19", 
                      yearid == "29" ~ "19-20")
         ) %>%
  mutate(yearid = as.character(yearid), 
       student_number = as.character(student_number), 
       studentid = as.character(studentid), 
       schoolid = as.character(schoolid), 
       grade_level = as.factor(grade_level), 
       att_date = ymd(att_date)
       ) %>%
  left_join(school_id_abbr, 
          by = "schoolid") %>%
  select(student_number,
          studentid,
          lastfirst,
          dob,
          gender,              
          grade_level,
          schoolid,
          abbr,
          att_date,
          yearid,
          sy,
          street,
          city,
          state,             
          zip,
          schoolentrydate,
          schoolentrygradelevel,
          enrolled,
          present,
          absent,
          tardy, 
         exitcode, 
         exitcomment
         ) %>%

  # Change empty strings to NA to more easily identify them
  mutate(
      street = na_if(street, ""),
      gender = na_if(gender, ""),
      city = na_if(city, ""),
      state = na_if(state, ""),
      zip = na_if(zip, "")
      )

```

The "student_identifying_info" table pulls together all identifying information for students in one place. This datframe is one row per student. 
```{r}
student_identifying_info <-
  student_daily_attendance %>%
  select(student_number, 
         lastfirst,
         dob,
         grade_level,
         gender,
         street,
         city,
         state,
         zip, 
         exitcode, 
         exitcomment) %>%
  distinct() %>%
  arrange(dob)
```

This dataset shows us the end of year average daily attendance for each student. Each row represents one students ada for an entire school year. The number of rows that a student appears in in the dataset corresponds to the number of years the student has been with KIPP Chiago (Note: We have removed the current 19-20 school year).
```{r}
student_yearly_ada <- 
  student_daily_attendance %>% 
  filter(yearid < CURRENT_FIRST_YEAR) %>%
  group_by(yearid,
           studentid, 
           student_number) %>% 
  summarize(enrolled = sum(enrolled),
            absent = sum(absent),
            year_end_atten_pct = 1 - (absent/enrolled))
```

The "student_cum_ada" gives a running total of each students average daily attendance each day for each year. For example, you could filter for yearid == 28 (last school year), and check each childs average daily attendance at a specific date. 
```{r}
student_cum_ada <- 
  student_daily_attendance %>%
  
  # only looking at students before the 19-20 school year
  filter(yearid < CURRENT_FIRST_YEAR) %>%
  group_by(yearid,
           studentid, 
           student_number) %>% 
  arrange(student_number, 
          att_date) %>%
  mutate(cum_enrolled = cumsum(enrolled),
         cum_absent = cumsum(absent),
         running_ada = 1 - (cum_absent/cum_enrolled)) %>%
  
  # only keep records for when students were actually enrolled. 
  filter(cum_enrolled > 0) %>%
  
  select(student_number, 
         studentid, 
         dob, 
         gender, 
         grade_level, 
         schoolid, 
         yearid,
         sy, 
         att_date, 
         present, 
         absent, 
         cum_present = cum_enrolled, 
         cum_absent, 
         running_ada
         )
```

## Exploratory Data Analysis

### Total Number of Students in the Dataset
```{r}
student_daily_attendance %>%
  select(student_number) %>%
  distinct() %>%
  count()
```

### Total number of students per school year
```{r}
student_daily_attendance %>%
  select(student_number, 
         sy) %>%
  distinct() %>%
  ggplot(aes(x = sy)) +
  geom_bar() +
  coord_flip() +
  labs(title="KIPP Chicago Student Count by School Year",
       x = "School Year", 
       y = "Number of Students")
```

### Student Numbers by Grade 18-19 School Year
```{r}
student_daily_attendance %>%
  filter(sy == "18-19") %>%
  select(student_number, 
         abbr, 
         grade_level) %>%
  distinct() %>%
  ggplot(aes(abbr)) +
  geom_bar(aes(fill=grade_level)) +
  labs(title="KIPP Chicago Student Count by School 18-19 SY",
       subtitle = "Schools have been subset by grade",
       x = "School Abbreviation",
       y = "Number of Students", 
       fill = "Grade Level")
```


## Explore Missingness
Dataset is missing around 1.3% of data. Mostly for students in the dob, gender and address categories. 
will search ASPEN to see if they can fill in this missing info. 
```{r}
vis_miss(student_identifying_info)
```
There are 123 students with missing personal data. 

```{r}
students_w_missing_ii_data <- 
  student_identifying_info %>% 
  filter(is.na(dob) | is.na(gender) | is.na(street) | is.na(city) | is.na(state))
```

### Find current students with missing identifying information
There are two current students with missing information. Student Number 50083869 and 50493973. I will check ASPEN to see if I can fill in the missing data by hand. 

The remaining 121 students have either transferred or have graduated and we no longer have a way to access prior information (such as address, dob or gender). These 121 students will be dropped from the dataset.
```{r}
current_student_check <- 
  students_w_missing_ii_data %>%
  filter(exitcode != 2 & exitcode != 14 & exitcode != 31 & exitcode != 32 & exitcode != 33)
```
## Next Steps

1. Figure out if there is a pattern in missing data